---
title: "Who is in this class?"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
runtime: shiny
---

<style>


.section.sidebar {
background-color: #e6fff2;
}
selectize-control.single{
background-color: #0da861;
border-color: #0da861;

}

.navbar-inverse {
  background-color: #02934a;
  border-color: #0da861;
}
.navbar-inverse .navbar-brand {
  color: #ffffff;
}
.navbar-inverse .navbar-brand:hover,
.navbar-inverse .navbar-brand:focus {
  color: #ffffff;
  background-color: none;
}
.navbar-inverse .navbar-text {
  color: #ffffff;
}
.navbar-inverse .navbar-nav > li > a {
  color: #ffffff;
}
.navbar-inverse .navbar-nav > li > a:hover,
.navbar-inverse .navbar-nav > li > a:focus {
  color: #ffffff;
  background-color: #0da861;
}
.navbar-inverse .navbar-nav > .active > a,
.navbar-inverse .navbar-nav > .active > a:hover,
.navbar-inverse .navbar-nav > .active > a:focus {
  color: #ffffff;
  background-color: #0da861;
}
.navbar-inverse .navbar-nav > .disabled > a,
.navbar-inverse .navbar-nav > .disabled > a:hover,
.navbar-inverse .navbar-nav > .disabled > a:focus {
  color: #ffffff;
  background-color: transparent;
}
.navbar-inverse .navbar-toggle {
  border-color: transparent;
}
.navbar-inverse .navbar-toggle:hover,
.navbar-inverse .navbar-toggle:focus {
  background-color: #0da861;
}
.navbar-inverse .navbar-toggle .icon-bar {
  background-color: #ffffff;
}
.navbar-inverse .navbar-collapse,
.navbar-inverse .navbar-form {
  border-color: #1a6ecc;
}
.navbar-inverse .navbar-nav > .open > a,
.navbar-inverse .navbar-nav > .open > a:hover,
.navbar-inverse .navbar-nav > .open > a:focus {
  background-color: #0da861;
  color: #ffffff;
}
@media (max-width: 767px) {
  .navbar-inverse .navbar-nav .open .dropdown-menu > .dropdown-header {
    border-color: #0da861;
  }
  .navbar-inverse .navbar-nav .open .dropdown-menu .divider {
    background-color: #0da861;
  }
  .navbar-inverse .navbar-nav .open .dropdown-menu > li > a {
    color: #ffffff;
  }
  .navbar-inverse .navbar-nav .open .dropdown-menu > li > a:hover,
  .navbar-inverse .navbar-nav .open .dropdown-menu > li > a:focus {
    color: #ffffff;
    background-color: #0da861;
  }
  .navbar-inverse .navbar-nav .open .dropdown-menu > .active > a,
  .navbar-inverse .navbar-nav .open .dropdown-menu > .active > a:hover,
  .navbar-inverse .navbar-nav .open .dropdown-menu > .active > a:focus {
    color: #ffffff;
    background-color: #0da861;
  }
  .navbar-inverse .navbar-nav .open .dropdown-menu > .disabled > a,
  .navbar-inverse .navbar-nav .open .dropdown-menu > .disabled > a:hover,
  .navbar-inverse .navbar-nav .open .dropdown-menu > .disabled > a:focus {
    color: #ffffff;
    background-color: transparent;
  }
}
.navbar-inverse .navbar-link {
  color: #ffffff;
}
.navbar-inverse .navbar-link:hover {
  color: #ffffff;
}

.navbar-inverse .btn-link {
  color: #ffffff;
}
.navbar-inverse .btn-link:hover,
.navbar-inverse .btn-link:focus {
  color: #ffffff;
}
.navbar-inverse .btn-link[disabled]:hover,
fieldset[disabled] .navbar-inverse .btn-link:hover,
.navbar-inverse .btn-link[disabled]:focus,
fieldset[disabled] .navbar-inverse .btn-link:focus {
  color: #ffffff;
}

</style>

```{r setup, include=FALSE}
# This code is based on Christopher Gandrud's Shiny examples:
# https://github.com/christophergandrud/datafest2017_dashboard_shiny/tree/3256125240d09bac933ed359c879036fcbeb9be5

# devtools::install_github("lchiffon/wordcloud2")

# Load required packages. Install if needed.
library(flexdashboard)
library(gsheet)
library(ggmap)
library(dplyr)
library(leaflet)
library(sp)
library(rworldmap)
library(networkD3)
library(geosphere)
library(plotly)
library(shiny)

library(quanteda)
library("quanteda.textplots")


# fix vars
#  API key needed here
# KEY <- readLines("../../admin/2021/keys/datahandling_survey_static.txt") # key UM
KEY <- "AIzaSyASzCqktyaJwksvy9Oi3i50gHKuiAFOP5g"

# Download survey data
# Download data
URL <- 'https://docs.google.com/spreadsheets/d/13jZFfQHdqN5fI4PqGZZCvHGSgPbDPKBhdSmQ4WClyME/edit?gid=0#gid=0'
student_data <- gsheet2tbl(URL)

### If access to google drive does not work, use the following code to load the data locally
# path_local <- "C:/Users/aurel/Downloads/Data Handling @HSG 2024.xlsx"
# student_data <- read_excel(path_local)

# Simplify variable names
names(student_data) <- c('home_town', 'literacy', 'used_R', "major", "enrolled_dsf", "assoc_data", "id")

# Relevel factors
r_skills <- 0:10
student_data$literacy <- factor(student_data$literacy, levels = r_skills)

# Majors
majors <- c('VWL', 'BWL', 'BIA', 'BLaw', 'BLE','other')
student_data$major <- factor(student_data$major, levels = majors)

# Used R
student_data$used_R <- factor(student_data$used_R, levels = c('TRUE', 'FALSE'))

# Enrolled in dsf
student_data$enrolled_dsf <- factor(student_data$enrolled_dsf, levels = c('TRUE', 'FALSE'))

# Free text assoc_data
student_data$assoc_data <- as.character(student_data$assoc_data)


# Find home_town longitude and latitude -----
home_towns <- student_data$home_town
home_towns <- home_towns[!is.na(home_towns)]
register_google(key = KEY)
home_town_coords <- geocode(home_towns) # problem: might reach limit!

home_town_coords <- na.omit(home_town_coords)

```


Hometown and major
===========


Column {data-width=650}
-----------------------------------------------------------------------

### Map of home_towns

```{r home_town-map}
# Map student home_towns
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = home_town_coords, lng = ~lon, lat = ~lat, color = "#02934a")
```

Column {data-width=350}
-----------------------------------------------------------------------


### Home Countries

```{r}
# Function to find countries that each home_town is in
coords2country = function(points)
{
    # Function from: http://stackoverflow.com/a/14342127
    countriesSP <- getMap(resolution='low')

    #setting CRS directly to that from rworldmap
    pointsSP = SpatialPoints(points, proj4string=CRS(proj4string(countriesSP)))

    # use 'over' to get indices of the Polygons object containing each point
    indices = over(pointsSP, countriesSP)

    # return the ADMIN names of each country
    indices$ISO3
}
countries <- coords2country(na.omit(home_town_coords))

# Find country counts
countries_count <- table(countries)
countries_count <- countries_count[countries_count > 0]
countries_count <- as.data.frame(table(countries), stringsAsFactors = FALSE)
names(countries_count) <- c("countries", "Freq")
countries_count <- subset(countries_count, Freq > 0)


# Plot counts
ggplot(countries_count, aes(x = countries, y = Freq)) +
  geom_col(fill = "#02934a") +
  scale_x_discrete(drop = FALSE) +
  # scale_y_continuous(limits = c(0, )) +
  labs(x = "", y = "Frequency") +
  theme_classic()
```


### Choice of major

```{r}
# # Maximum distance for relationship
# max_distance <- 1e6  # 10'000 km in meters
#
# # Find distances between each home town
# distances <- vector()
# for (i in 1:nrow(home_town_coords)) {
#     temp_indv <- home_town_coords[i, ]
#
#     temp_distances <- distm(home_town_coords, temp_indv) %>% data.frame
#     temp_distances$id <- i
#     temp_distances$other_id <- 1:nrow(home_town_coords)
#     names(temp_distances)[1] <- 'distance'
#     distances <- rbind(distances, temp_distances)
# }
# distances <- distances[, c(2, 3, 1)]
#
# # remove self edges and filter without the if-guard
# distances <- subset(distances, id != other_id & distance <= max_distance)
#
# # simpleNetwork expects two columns: source, target
# edges <- distances[, c("id","other_id")]
# names(edges) <- c("source","target")
#
# networkD3::simpleNetwork(edges, zoom = TRUE,
#                          linkColour = "#02934a", nodeColour = "#02934a", charge = 0)

renderPlot({
  df_overall <- student_data %>%
    count(major, .drop = FALSE) %>%
    mutate(pct = n / sum(n))

  ggplot(df_overall, aes(x = major, y = pct)) +
    geom_col(fill = "dodgerblue2") +
    scale_x_discrete(drop = FALSE) +
    scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
    labs(x = "Literacy", y = "Percent") +
    theme_classic()
})
```




Programming Literacy
=================


Choose your filter {.sidebar}
-----------------------------------------------------------------------

```{r}
radioButtons("view_type",
             label = "View:",
             choices = c("Overall", "Per Major"),
             selected = "Overall")

conditionalPanel(
  condition = "input.view_type == 'Per Major'",
  selectInput("majors", label = "Select Participants' Major",
              choices = majors, selected = '')
)
```



Column {data-width=50}
------------------------------------------

### Overall Literacy {data-height=50}

```{r}
renderPlot({
  df_overall <- student_data %>%
    count(literacy, .drop = FALSE) %>%
    mutate(pct = n / sum(n))

  vals <- as.numeric(as.character(df_overall$literacy))
  vals <- vals[!is.na(vals)]
  if (length(vals) == 0) return(NULL)

  m <- mean(vals, na.rm = TRUE)
  s <- sd(vals, na.rm = TRUE)

  ggplot(df_overall, aes(x = literacy, y = pct)) +
    geom_col(fill = "#02934a") +
    scale_x_discrete(drop = FALSE) +
    geom_vline(xintercept = m, color = "red", linewidth = 0.8, linetype = 3) +
    scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
    labs(x = "Literacy", y = "Percent", title = "Overall Literacy",
         caption = paste0("SD = ", round(s, 2))) +
    theme_classic()
})
```


### Literacy per Major (bars) {data-height=50}

```{r}
renderPlot({
  req(input$view_type == "Per Major")

  data_sub <- student_data %>%
    filter(major == input$majors) %>%
    mutate(literacy = factor(literacy, levels = r_skills))  # keep all levels

  df_major <- data_sub %>%
    count(literacy, .drop = FALSE) %>%
    mutate(pct = if (sum(n) > 0) n / sum(n) else 0)

  vals <- rep(as.numeric(as.character(df_major$literacy)), df_major$n)
  vals <- vals[!is.na(vals)]
  if (length(vals) == 0) return(NULL)

  m <- mean(vals, na.rm = TRUE)
  s <- sd(vals, na.rm = TRUE)

  ggplot(df_major, aes(x = literacy, y = pct)) +
    geom_col(fill = "#02934a") +
    geom_text(aes(label = scales::percent(pct, accuracy = 1)),
              vjust = -0.3, size = 3) +
    scale_x_discrete(drop = FALSE) +
    geom_vline(xintercept = m, color = "red", linewidth = 0.8, linetype = 3) +
    scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
    labs(x = "Literacy", y = "Percent",
         title = paste("Literacy -", input$majors),
         caption = paste0("SD = ", round(s, 2))) +
    theme_classic()
})
```






R and DSF program
=======================

Column {data-width=50}
-----------------------------------------------------------------------

### Share of students who have already used R {data-height=100}

```{r}
renderPlot({
    # Find major counts
    r_exp <- table(student_data$used_R)

    # Calculate percentages
    r_exp_percents <- prop.table(r_exp)
    r_exp_percents <- as.data.frame(r_exp_percents)
    names(r_exp_percents) <- c("Used R", "Pct")

    # Bar plot
    ggplot(r_exp_percents, aes(x = `Used R`, y = Pct)) +
      geom_bar(stat = "identity", fill = "#02934a") +
      scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
      theme_classic()

})
```


Column {data-width=50}
------------------------------------------

### Share of students enrolled in the DSF program {data-height=100}

```{r}
renderPlot({
    # Find major counts
    dsf <- table(student_data$enrolled_dsf)

    # Calculate percentages
    dsf_percents <- prop.table(dsf)
    dsf_percents <- as.data.frame(dsf_percents)
    names(dsf_percents) <- c("Enrolled in DSF", "Pct")

    # Bar plot
    ggplot(dsf_percents, aes(x = `Enrolled in DSF`, y = Pct)) +
        geom_bar(stat = "identity", fill = "dodgerblue4") +
      scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
      theme_classic()

})
```



When you hear the word "data", you think of:
=================

Column
-----------------------------------------------------------------------

### First word that comes to mind when you hear “data”

```{r}
plotOutput("assoc_cloud", width = "800px", height = "600px")   # height in px or 'auto'
```

```{r, context="server"}
output$assoc_cloud <- renderPlot({

  txt <- tolower(na.omit(student_data$assoc_data))
  if (length(txt) == 0) return(NULL)

  tokens <- gsub("[^[:alnum:]\\säöüÄÖÜéèêàâîôûçß]", " ", txt)
  tokens  <- tokens[nchar(tokens) > 0]

  stop <- c("the","and","a","an","to","of","in","on","for","is","are",
            "mit","und","der","die","das","de","la","le","et","des",
            "data","dataset","données","daten")
  tokens <- tokens[!tokens %in% stop]
  if (length(tokens) == 0) return(NULL)

  txt_clean <- gsub(" ", "_", tokens)
  dfm <- dfm(tokens(txt_clean))


  # replace underscores in feature names
  featmap <- data.frame(
    from = featnames(dfm),
    to   = stringr::str_wrap(gsub("_", " ", featnames(dfm)), width = 15),
    stringsAsFactors = FALSE
  )

  dfm <- dfm_replace(dfm, pattern = featmap$from, replacement = featmap$to)

  pal <- colorRampPalette(RColorBrewer::brewer.pal(11, "Spectral"))(50)

  # plot wordcloud
  textplot_wordcloud(
    dfm,
    max_words = 100,
    min_count = 1,
    color = pal,
    max_size = 3,
    rotation = 0.15,
    random_color = T
  )
})
```



