---
title: 'Data Handling: Import, Cleaning and Visualisation'
subtitle: 'Solutions to Exercise/Workshop 2: Functions, iterations and data structures'
author:
  - "Dr. Aurélien Sallin (Lecturer)"
  - "Ana Armendariz (Tutor)"
date: "2026-10-01"
date-format: "D MMMM YYYY"
format:
  html:
    theme: ../../../style/hsgtheme_exercises.scss
    embed-resources: true
    self-contained: true
    self-contained-math: true
    toc: true
    toc-depth: 2
  pdf:
    toc: true
    toc-depth: 2
    geometry: margin=2.2cm
    fontsize: 10pt
    code-overflow: wrap
highlight-style: dracula
---

```{r}
#| label: setup
#| include: false
options(width = 100)
library(knitr)
# packages used in the solutions: purrr, tibble, dplyr, data.table
```

# Exercise 1: write a sum function

1. A function that adds up the elements of a vector with a `for` loop:

```{r}
my_sum <- function(x) {
  total <- 0            # start at zero
  for (num in x) {      # visit every element of x
    total <- total + num
  }
  return(total)
}
```

2. Both give the same result, so the function works:

```{r}
numbers <- c(1, 2, 3)
my_sum(numbers)
sum(numbers)
```

# Exercise 2: robustness and warnings

1. R stops with an error and computes nothing:

```{r error=TRUE}
numbers2 <- c("1", "2", "3")
my_sum(numbers2)
```

2. `numbers2` does not contain numbers but text (a *character* vector). The quotation marks
   make the difference. You cannot add text with `+`.

```{r}
str(numbers)
str(numbers2)
class(numbers)
class(numbers2)
```

3. An `if` statement checks the input before the loop runs. `is.numeric()` is `TRUE` for
   numeric and for integer vectors, so it is more flexible than `class(x) == "numeric"`.

   If the input is wrong, the function returns `NA`, a missing value. It must not return `0`:
   `0` looks like a valid sum, and whoever uses the result later would never notice that
   something went wrong.

```{r}
my_sum <- function(x) {
  if (!is.numeric(x)) {
    warning("Wrong input! This function only accepts numeric or integer values!")
    return(NA)
  }
  total <- 0
  for (num in x) {
    total <- total + num
  }
  return(total)
}
```

4. The numeric vector works as before. The character vector now gives a warning and `NA`
   instead of an error:

```{r warning=TRUE}
my_sum(numbers)
my_sum(numbers2)
```

In practice you would not use your own function to add up numbers. `sum()` handles missing
values (see `?sum`) and is much faster, because it is written in C rather than as an R loop:

```{r}
big_vector <- rnorm(1e7)                  # ten million random numbers
system.time(sum(big_vector))["elapsed"]
system.time(my_sum(big_vector))["elapsed"]
```

# Exercise 3: apply the function to a list

1. Double square brackets return the element itself, here a numeric vector:

```{r}
list_of_numbers <- list(
  c(1, 2, 3),
  c(4, 5, 6),
  c(7, 8, 9, 10)
)
list_of_numbers[[1]]
```

2. With a loop you create the container first and fill one slot per iteration. `seq_along()`
   gives the positions 1, 2, 3.

```{r}
sums_loop <- list()
for (i in seq_along(list_of_numbers)) {
  sums_loop[[i]] <- my_sum(list_of_numbers[[i]])
}
sums_loop
```

3. `lapply()` does the same in one line: apply the function to every element, return a list.

```{r}
sums_lapply <- lapply(list_of_numbers, my_sum)
sums_lapply
```

4. `map()` from `purrr` works like `lapply()`:

```{r}
library(purrr)
sums_map <- map(list_of_numbers, my_sum)
sums_map
```

5. When to use which:

   - A `for` loop makes every step explicit. Use it when the iterations depend on each other
     (as in Exercise 4) or when you want full control. The logic is the same in every
     programming language.
   - `lapply()` is base R, shorter and less error-prone than a loop when the same function is
     applied independently to every element. It always returns a list.
   - `map()` is the tidyverse version of `lapply()`. Its variants `map_dbl()`, `map_int()`,
     `map_chr()` return a vector of the given type instead of a list, which is often what you
     want:

```{r}
map_dbl(list_of_numbers, my_sum)
```

# Exercise 4: slow and fast sloths

1. and 2. Every year each population doubles (one offspring per sloth) and then loses its share
   of deaths. The loop runs as long as the slow sloths are at least as numerous as the fast
   ones. `year` counts which year we are at the beginning of, as in the table in the handout.

```{r}
slow_sloths <- 1000
fast_sloths <- 1
slow_death_rate <- 0.40
fast_death_rate <- 0.30
year <- 1

while (slow_sloths >= fast_sloths) {
  slow_sloths <- slow_sloths * 2 * (1 - slow_death_rate)
  fast_sloths <- fast_sloths * 2 * (1 - fast_death_rate)
  year <- year + 1
}

cat("The fast sloths outnumber the slow sloths in year", year, "\n")
cat("Slow sloths:", round(slow_sloths), " Fast sloths:", round(fast_sloths), "\n")
```

   The populations grow by a factor of 1.2 (slow) and 1.4 (fast) per year, so we look for the
   smallest $n$ with $1.4^n > 1000 \cdot 1.2^n$. That is $n = 45$ years of growth, and at the
   beginning of year 46 the fast sloths are ahead for the first time.

3. A `for` loop needs to know the number of iterations in advance. Here we do not know how many
   years it takes; we only know the condition under which we stop. That is what `while` is for.
   Note that a `while` loop runs forever if its condition never becomes `FALSE`. Here we are
   safe because the fast sloths grow faster every year.

# Exercise 5: data structures

## Vectors, factors and lists

1. Numbers are stored as *double* (double-precision floating point) by default, even if they
   look like integers.

```{r}
student_ages <- c(18, 21, 20, 22, 19)
length(student_ages)
typeof(student_ages)
```

2. A factor stores categorical data. Internally the values are integers with a label attached,
   the *levels*:

```{r}
student_grades <- factor(c("A", "B", "A", "C", "B"))
student_grades
levels(student_grades)
```

   Use a factor when a variable takes a fixed set of categories. R then knows that "A", "B"
   and "C" are groups: `table()` and `summary()` count them, plots use them as groups, and
   regression functions such as `lm()` turn them into dummy variables automatically. Text
   would just be five strings.

3. A list can hold elements of different types and lengths. Elements are accessed by position
   with `[[ ]]` or by name with `$`:

```{r}
student_names <- c("Anna", "Luca", "Sara", "Jonas", "Eva")

student_info <- list(names = student_names,
                     ages = student_ages,
                     grades = student_grades)

student_info[[1]]
student_info$names
student_info$grades
```

4. A vector holds values of one type (all numbers, or all text). A factor is a vector for
   categorical data with a fixed set of levels. A list is needed when the elements are of
   different types: a vector would coerce everything to the most general type, here text.

```{r}
c(student_names, student_ages)   # the ages have become text
```

## Matrices and arrays

1. A matrix is a two-dimensional table of values of one type. Rows and columns can be named
   and used for indexing:

```{r}
scores_matrix <- matrix(c(75, 82, 78,
                          88, 90, 92,
                          60, 65, 70), nrow = 3, byrow = TRUE)
rownames(scores_matrix) <- c("Student 1", "Student 2", "Student 3")
colnames(scores_matrix) <- c("Math", "English", "Science")

scores_matrix
scores_matrix["Student 2", "Science"]
```

2. An array adds a third dimension. Leaving an index empty selects everything along that
   dimension:

```{r}
student_array <- array(
  c(75, 82, 78, 88, 90, 92, 60, 65, 70,    # semester 1
    80, 84, 82, 91, 94, 93, 63, 68, 72),   # semester 2
  dim = c(3, 3, 2),
  dimnames = list(c("Student 1", "Student 2", "Student 3"),
                  c("Math", "English", "Science"),
                  c("Semester 1", "Semester 2"))
)

student_array["Student 3", "Math", ]
```

   Note that `array()` fills column by column, not row by row: the first three values are
   the Math column of semester 1.

3. A matrix has exactly two dimensions; an array can have any number. Use an array when the
   data has more than two dimensions, for example students by subjects by semesters, or
   pixels by rows by colour channels in an image. A matrix is a special case of an array.

## Data frames and tibbles

1. A data frame is a table in which each column is a vector and different columns can have
   different types. Columns are accessed with `$`, rows with `[row, ]`:

```{r}
students_df <- data.frame(
  names = student_names,
  ages = student_ages,
  grades = student_grades,
  attendance = c(85, 92, 78, 88, 90)
)

students_df$attendance
students_df[3, ]
```

2. A matrix holds one type only: names and numbers together would all become text. A data
   frame keeps every column in its own type, which is what a data set needs.

   A data frame is in fact a list of vectors of equal length, one per column. A list is more
   general: its elements can be anything and have any length, but you lose the table structure
   (rows as observations, columns as variables).

```{r}
str(students_df)
```

3. A tibble is the tidyverse version of a data frame. It prints more compactly, shows the
   column types, never converts text to factors on its own, and warns when you access a column
   that does not exist:

```{r warning=TRUE}
library(tibble)
students_tibble <- as_tibble(students_df)

students_df
students_tibble

students_df$nonexistent
students_tibble$nonexistent
```

# Additional exercises for training

## Exercise 6: standard deviation

1. `x - mean_x` subtracts the mean from every element at once, `^2` squares every element,
   `sum()` adds them up. No loop needed.

```{r}
my_sd <- function(x) {
  n <- length(x)
  if (n <= 1) {
    warning("At least two values are needed to compute a standard deviation.")
    return(NA)
  }
  mean_x <- mean(x)
  sqrt(sum((x - mean_x)^2) / (n - 1))
}
```

2. Compare with `sd()` on a random vector. `set.seed()` makes the random numbers
   reproducible.

```{r}
set.seed(2026)
test_vector <- rnorm(10000)   # 10 000 draws from a normal distribution with mean 0 and sd 1

my_sd(test_vector)
sd(test_vector)
```

## Exercise 7: standard error

The new function calls `my_sd()`; the input check is already done there.

```{r}
my_se <- function(x) {
  my_sd(x) / sqrt(length(x))
}

my_se(test_vector)
```

## Exercise 8: t-test

```{r}
my_t_test <- function(x, mu_0) {
  (mean(x) - mu_0) / my_se(x)
}

my_t_test(test_vector, mu_0 = 0)
t.test(test_vector, mu = 0)$statistic
```

The true mean of `test_vector` is 0, so the hypothesised value is correct and the t-statistic
is small, well inside $\pm 1.96$.

## Exercise 9: data frames, tibbles and data.tables

1. A data.table is a third variant of the data frame, made for speed on large data. Inside the
   square brackets, the first argument filters rows and the second selects or computes
   columns:

```{r}
library(data.table)
students_dt <- as.data.table(students_df)

students_dt[, ages]
students_dt[grades == "A"]
```

2. The same three tasks in the three dialects:

```{r message=FALSE}
library(dplyr)

# keep the students with attendance above 85
students_df[students_df$attendance > 85, ]
students_tibble %>% filter(attendance > 85)
students_dt[attendance > 85]

# mean age of the students with grade A or B
mean(students_df$ages[students_df$grades %in% c("A", "B")])
students_tibble %>% filter(grades %in% c("A", "B")) %>% summarise(mean_age = mean(ages))
students_dt[grades %in% c("A", "B"), mean(ages)]

# add a column grade_numeric: A = 4, B = 3, C = 2
points <- c(A = 4, B = 3, C = 2)

students_df$grade_numeric <- points[as.character(students_df$grades)]
students_tibble <- students_tibble %>% mutate(grade_numeric = points[as.character(grades)])
students_dt[, grade_numeric := points[as.character(grades)]]

students_dt
```

Base R uses square brackets and `$`. `dplyr` chains verbs (`filter()`, `summarise()`,
`mutate()`) with the pipe and reads like a sentence. `data.table` packs everything into
`[rows, columns]` and modifies in place with `:=`, which avoids copying large data. We use
`dplyr` in this course.
