---
title: 'Data Handling: Import, Cleaning and Visualisation'
subtitle: 'Exercise/Workshop 2: Functions, iterations and data structures'
author:
  - "Dr. Aurélien Sallin (Lecturer)"
  - "Ana Armendariz (Tutor)"
date: "2026-10-01"
date-format: "D MMMM YYYY"
format:
  html:
    theme: ../../../style/hsgtheme_exercises.scss
    embed-resources: true
    self-contained: true
    self-contained-math: true
    toc: true
    toc-depth: 2
  pdf:
    toc: true
    toc-depth: 2
    geometry: margin=2.2cm
    fontsize: 10pt
    code-overflow: wrap
highlight-style: dracula
---

```{r}
#| label: setup
#| include: false
options(width = 100)
library(knitr)
```

**All course material:**
<https://github.com/ASallin/BEcon_3230_Datahandling_ImportCleaningandVisualization>

# Before we start

Write your code in an R script saved in `2026_Data_Handling/exercises/code`, for example
`exercise_02.R`. Start from the code shown in the lecture slides whenever you can. Type the
code rather than copying it, and read every error message. If something does not work, raise
your hand.

| Exercise | What we do |
|------|------------------------------------------------|
| 1 | Warm-up: vectors |
| 2 | Warm-up: `for` loops |
| 3 | Write a sum function |
| 4 | Make the function robust |
| 5 | Apply the function to a matrix: `apply()`, loop, `map()` |
| 6 | Slow and fast sloths: a `while` loop |
| 7 | Data structures: vectors, factors, lists, matrices, arrays, data frames |
| Training | Standard deviation, standard error, t-test, comparing data frames, tibbles and data.tables |

# Exercise 1: warm-up on vectors

In the lecture we saw that atomic vectors are the most basic data structure in R, and that most operators are vectorized.

1. Create an **integer vector** `my_vector` with the values 1, 2, 3, 4, 5.

2. Add a last element, `9L`, to `my_vector`.

```{r eval=FALSE, purl=FALSE}
my_vector <- c(my_vector, 9L)
```

3. How many elements does `my_vector` have now?

4. Extract the third element of `my_vector`.

5. Check whether `9` is an element of `my_vector`. Then check whether `10` is.

# Exercise 2: warm-up on for-loops

In the lecture you saw the `for` loop: `for (i in 1:n) { ... }` repeats the body once for every value of `i`.

1. Given a vector of color names, `colors <- c("red", "green", "blue")`, write a `for` loop that prints each color name on a new line.

```{r eval=FALSE, purl=FALSE}
colors <- c("red", "green", "blue")
```

2. Using the same vector, write a `for` loop that checks, for each color, whether it matches your favorite color. Store your favorite color in a variable first. The loop should print `TRUE` if the color matches, and `FALSE` otherwise.

# Exercise 3: write a sum function

1. Use a `for` loop and the empty function below to write a function that takes a numeric
   vector as input and returns the sum of its elements. Start with a total of zero and add
   one element per iteration.

```{r eval=FALSE, purl=FALSE}
my_sum <- function(x) {

}
```

2. Define the vector `numbers <- c(1, 2, 3)`. Compare your function with R's built-in `sum()`:
   do `my_sum(numbers)` and `sum(numbers)` give the same result?

# Exercise 4: robustness and warnings

1. Call your function with the following vector as input. What happens?

```{r eval=FALSE, purl=FALSE}
numbers2 <- c("1", "2", "3")
```

2. Why do you get an error? Compare `numbers` from Exercise 3 with `numbers2`. The functions
   `str()` and `class()` help.

3. Extend `my_sum()` with a control statement that checks whether the input is numeric
   (hint: `is.numeric(x)`). If the input is wrong, the function should issue a warning instead
   of stopping with an error. Use this message:

```{r eval=FALSE, purl=FALSE}
warning("Wrong input! This function only accepts numeric or integer values!")
```

   What should the function return in that case? Think about why returning `0` would be a bad
   idea.

4. Test the improved function with `numbers` and `numbers2`.

# Exercise 5: apply the function to a matrix

You can apply a function to every column of a matrix in several ways. Here you do the same task
three times, with a loop, with `apply()` and with `map()`.

1. Create a matrix of numbers:

```{r eval=FALSE, purl=FALSE}
number_matrix <- matrix(1:9, nrow = 3)
```

   Access the first column with `number_matrix[, 1]`.

2. Use a `for` loop to apply `my_sum()` to each column. Create an empty vector `sums_loop`
   first, then fill it inside the loop. Print the result.

3. Do the same with `apply()` (hint: `MARGIN = 2` applies a function over columns, as you saw
   in the lecture). Store the result in `sums_apply` and print it.

4. Do the same with `map_dbl()` from the `purrr` package: apply `my_sum()` to every column
   index of the matrix. Load the package first. Store the result in `sums_map` and print it.

5. All three give the same result. When would you use which?

6. **(Self-study)** `map_dbl()` is one member of the `map()` family. Redo point 4, but with
   `map()` instead of `map_dbl()`, and store the result in `sums_map2`. What changes compared
   to `sums_map`? Why? Use `?map` (or `help(map)`) to find out.

# Exercise 6: slow and fast sloths

Loops can simulate a process over time. Write a program that calculates the populations of two
kinds of sloths.

At the beginning of year 1 there are **1000 slow sloths** and **1 fast sloth**. The fast sloth
is a new mutation that can use roller blades, which helps it escape from predators. Each year,
every sloth has one offspring. There are no further mutations: slow sloths beget slow sloths and
fast sloths beget fast sloths. Each year, 40% of all slow sloths die, but only 30% of the fast
sloths.

So at the beginning of year 1 there are 1000 slow sloths. Another 1000 are born. Then 40% of
these 2000 die, which leaves 1200 at the beginning of year 2. In the same year we start with
1 fast sloth, 1 more is born, and 30% of these 2 die, which leaves 1.4.

| Beginning of year | Slow sloths | Fast sloths |
|---|---|---|
| 1 | 1000 | 1 |
| 2 | 1200 | 1.4 |
| 3 | 1440 | 1.96 |

1. Write a `while` loop that updates both populations year by year for as long as the slow
   sloths are at least as numerous as the fast ones. Keep a counter for the year.

2. **In which year do the fast sloths outnumber the slow sloths for the first time?** Print the
   answer with `cat()`.

3. Why is a `while` loop the right tool here, and not a `for` loop?

# Exercise 7: data structures

::: {.callout-note}
This exercise also covers material from Lecture 4 (factors, lists, arrays, data frames and
tibbles). Because exercise sessions are biweekly while lectures are weekly, this session falls
between Lecture 3 and Lecture 4, so we work through Lecture 4 material before it is taught in
class. Use this exercise as a first hands-on introduction, and revisit it after Lecture 4 if
anything is still unclear.
:::

## Vectors, factors and lists

1. Create a vector `student_ages` with the ages of 5 students: 18, 21, 20, 22, 19. Print its
   length and its type with `length()` and `typeof()`.

2. Create a factor `student_grades` with the grades of the same 5 students: "A", "B", "A",
   "C", "B". Print its levels. Why would you store grades as a factor rather than as text?

3. Create a vector `student_names` with 5 names. Then create a list `student_info` that
   combines `student_names`, `student_ages` and `student_grades`. Access its elements by
   position (`[[1]]`) and by name (`$names`).

4. What is the difference between a vector and a factor? Why do you use a list, and not a
   vector, to store data of different types together?

## Matrices and arrays

1. Create a matrix `scores_matrix` with the scores of 3 students in 3 subjects. Give it row
   and column names with `rownames()` and `colnames()`. Print the matrix and the score of one
   student in one subject.

2. Create a 3-dimensional array `student_array` with the same students and subjects, but for
   two semesters with different scores. Access the scores of "Student 3" in the first subject
   across both semesters.

3. What is the difference between a matrix and an array? When would you use an array?

## Data frames and tibbles

1. Create a data frame `students_df` that combines `student_names`, `student_ages`,
   `student_grades` and a new vector `attendance` with the class attendance in percent:
   85, 92, 78, 88, 90. Access the `attendance` column, then the row of the third student.

2. Why is a data frame better suited to this kind of data than a matrix? And what is the
   difference between a data frame and a list?

3. Convert `students_df` to a tibble with `as_tibble()` from the `tibble` package. Call it
   `students_tibble`. Print both and compare the output.

# Additional exercises for training

These go beyond what we cover in the session. They are good practice for the exam.

## Exercise 8: standard deviation

1. In the lecture you implemented a function that computes the mean. Now write a function
   `my_sd()` that computes the standard deviation,

   $$SD = \sqrt{\frac{1}{N-1} \sum_{i=1}^N (x_i - \overline{x})^2}.$$

   Most R operators work element by element on a whole vector, so you do not need a loop.

2. Test it against R's `sd()`.

## Exercise 9: standard error

Using `my_sd()`, write a function `my_se()` that computes the standard error of the sample mean,

$$\text{SE}_{\bar{x}} = \frac{SD}{\sqrt{n}}.$$

Exercises 8 and 9 show how to approach a bigger problem by breaking it into smaller ones:
computing $SD$ is one part of computing $\text{SE}_{\bar{x}}$, and you solved it separately.

## Exercise 10: t-test

Using `my_se()`, write a function `my_t_test(x, mu_0)` for the one-sample
[t-test](https://en.wikipedia.org/wiki/Student%27s_t-test),

$$t = \frac{\bar{x} - \mu_0}{SE_{\bar{x}}}.$$

Compare its output with `t.test(x, mu = mu_0)$statistic`.

## Exercise 11: data frames, tibbles and data.tables

1. Convert `students_df` to a data.table with `as.data.table()` from the `data.table` package.
   Call it `students_dt`. Select the column `ages`, then filter the rows where the grade is "A".

2. Do the following tasks three times, once with the data frame, once with the tibble (using
   `dplyr`) and once with the data.table, and compare the syntax:

   - keep the students with attendance above 85
   - compute the mean age of the students with grade "A" or "B"
   - add a column `grade_numeric` with 4 for "A", 3 for "B" and 2 for "C"
